Kindly Bent to Free Us

### 2019-07-07 kindly/infer/infer.tex
Fix missing refs.
Small fixes.

We now formulate type inference for \lang. Type inference is
based on the HM(X) framework~\citep{DBLP:journals/tapos/OderskySW99} which
presents how to implement Hindley-Milner type inference in a language
with qualified types and an arbitrary custom constraint language.
We adapt and extend the rules of HM(X) to support kind inference,
track linearity and affinity and handle borrow and regions. Furthermore, we
formulate an appropriate constraint solving algorithm for our constraint
system such that inference is sound and complete.
In the rest of this section, we first present various preliminaries definitions
before showing our kind and type inference judgements
and our constraint solving algorithm.

@@

The HM(X) framework assumes the existence of a $\operatorname{normalize}$
function which takes a constraint $C$ and a substitution $\psi$ and returns a
constraint in solved form $C'$,
and an updated substitution $\unif'$.
Normalization is supposed to return the ``best'' solution $(C',\unif')$, for
instance the most general unifier.
We detail the implementation
of the normalization function in \cref{infer:solving}. In the meantime, we just
assume the existence of such a function for our constraint system.

@@

\subsection{Constraint solving}
\label{infer:normalize}

>>>>

The main innovation of \affe is its principal type inference.
We now formulate our inference algorithm
based on the HM(X) framework~\citep{DBLP:journals/tapos/OderskySW99}.
HM(X) presents how to implement Hindley-Milner type inference in a language
with qualified types and an arbitrary custom constraint language.
We adapt and extend the rules of HM(X) to support kind inference,
track linearity and handle borrows and regions. Furthermore, we
formulate an appropriate constraint solving algorithm for our constraint
system such that inference is sound and complete.
We first present various preliminary definitions
before showing our kind and type inference judgements
and our constraint solving algorithm.

@@

The HM(X) framework assumes the existence of a $\operatorname{normalize}$
function which takes a constraint $C$ and a substitution $\psi$ and returns a
simplified constraint $C'$,
and an updated substitution $\unif'$.
Normalization is supposed to return the ``best'' solution $(C',\unif')$, for
instance the most general unifier.
We detail the implementation
of the normalization function and its property in \cref{infer:solving}.
In the meantime, we simply
assume the existence of such a function for our constraint system.

@@

\subsection{Constraint solving}
\label{infer:solving}

### 2019-07-07 kindly/motivation.tex
Typo fixes in the motivation.
Small fixes.

Line~\ref{line:session1} introduces a parameterized abstract
type \lstinline{st} which is linear as indicated
by its kind \lstinline{lin}. Its low-level implementation would wrap a
handle for a socket, for example. The \lstinline{receive}  operation
in Line~\ref{line:session2} takes a channel that is ready to receive a
value of type \lstinline{'a} and returns a pair of the value a the
channel at its residual type \lstinline{'S}. It does not matter
whether \lstinline{'a} is restricted to be linear, in fact
\lstinline{receive} is polymorphic in the kind of \lstinline{'a} and
also in the kind of \lstinline{->}. This kind polymorphism is the
default if no constraints are specified.
%
The \lstinline{send} operation takes a linear channel and returns a
single-use function that takes a value of type \lstinline{'a} suitable
for sending and returns the channel with updated type.
%
The \lstinline{create} operation returns a pair of channel
endpoints. They follow dual communication protocols, where the
\lstinline{dual} operator swaps sending and receiving operations.
%
Finally, \lstinline{close} closes the channel.

@@

\cref{ex:array:region} uses braces as in \lstinline/{| ... |}/ to
explicitly state the regions of the example in \cref{ex:array}.  One
region consists of the header expression of the \lstinline/let/ in
\cref{ex:array:region:get}. It is contained in another region that
comprises the body of the \lstinline/for/ loop. \lang guarantees that borrows 
never escape the smallest enclosing region. It employs a system of
\emph{indexed kinds} like \lstinline{aff_r} and \lstinline{un_r} where
$r$ is a positive integer that corresponds to the lexical nesting
depth of regions. For instance, the type of \lstinline{&!a} in
\cref{ex:array:set} has kind \lstinline{aff_1} whereas the type of
\lstinline{&a} in \cref{ex:array:region:get} has kind
\lstinline{unr_2} and the typing of its region is such that types with
kind indexes greater than or equal to $2$ cannot escape.

@@

\cref{sig:hybarray} contains an API \lstinline/HybArray/ 
that enables using mutable and immutable modifications to
the board through affine types and borrows.
The signature differs slightly from the \lstinline/Array/
signature. For simplicity, the array elements are unrestricted, but
the structure itself remains linear so as to be implemented in terms of \lstinline/Array/.
The in-place mutation function \lstinline/set_mut/ with type
\lstinline/&!('a t) * int * 'a -> unit/ works on a mutable borrow  whereas the persistent
\lstinline/set/ operation has type \lstinline/&('a t) * int * 'a -> 'a t/. It
takes an immutable borrow because it only reads from the
argument array, but returns a fresh,  modified structure.
The code in \cref{ex:cow} contains a very simple implementation of
\lstinline/HybArray/ that represents hybrid arrays
as regular arrays and uses copy-on-write for persistent
modifications. The function
\lstinline/create_from : int -> (int -> 'a) -> 'a Array.t /
initializes an array from a function.
% A more elaborate implementation could use persistent arrays
% \cite{DBLP:conf/ml/ConchonF07}.

@@

  let set (a, i, x) =
    let f k =
      if k = i then x else get (a,k)
    in
    create_from (length a) f

@@

The board is represented as a 2D-matrix (\cref{line:boardtype}), where
the \lstinline/Matrix/ type uses the same API as \lstinline/HybArray/
but with two indices. 
Each cell contains an integer set that represent admissible solutions so far.
The Sudoku solver iterates over the cells and tries each possible solution (\cref{line:try_solution}).
When a value is picked for the current cell, we create a choicepoint,
change the cell with an immutable modification (\cref{line:immutset}), and propagate
the changes with the \lstinline/propagate/ function.
The \lstinline/propagate/ function uses direct mutation through a
mutable borrow of the matrix as it need not preserve the previous
version of the board.  
The implementation of \lstinline/propagate/ is split into three parts
for lines, columns, and square, which are all very similar to function
\lstinline/propagate_lines/ (\cref{line:propline}).

The typing ensures that the mutations do not compromise the state at
the choice point, because they operate on a new state \lstinline/new_g/ created for one
particular branch of the choice.
In particular, the \lstinline/try_solution/ function can be used in a loop
precisely because the \lstinline/set/ function only requires an
unrestricted immutable borrow of \lstinline/g/, which can be used as
many times as needed.
As \lstinline/g/ is linear we must free it in \cref{line:free:g}.

@@

The implementation in \cref{impl:pool} represents a bag of resourfces
using a concurrent queue with atomic add and remove operations. 
The implementation of the \lstinline/Pool.create/ and \lstinline/Pool.consume/
functions is straightforward.
The function \lstinline/Pool.use/ first draws
an element from the pool (or creates a fresh element),
passes it down to the callback function \lstinline/f/, and returns
it to the pool afterwards.
We explicitly delimit the region in \cref{line:pool:region} to ensure that
the return value of \lstinline/f &!o/ does not capture \lstinline/&!o/.

\begin{figure}[tp]
  \centering
  \begin{subfigure}[t]{.55\linewidth}
\begin{lstlisting}
type ('a:'k) pool : 'k

@@

let uri = "postgresql://localhost:5432"
let main users =
  (* Create a database connection pool. *)
  let pool = Pool.create (fun _ -> Db.connect uri)
  List.parallel_iter (create_user &pool) users
  Pool.consume (fun c -> Db.close c)
\end{lstlisting}
    \caption{Example of use}
    \label{ex:pool}
  \end{subfigure}~
  \begin{subfigure}[t]{.45\linewidth}
\begin{lstlisting}
type ('a:'k) pool : 'k = {
  spawn : unit -> 'a ;
  queue : 'a CQueue.t;
}

>>>>

Line~\ref{line:session1} introduces a parameterized abstract
type \lstinline{st} which is linear as indicated
by its kind \lstinline{lin}. Its low-level implementation would wrap a
handle for a socket, for example. The \lstinline{receive}  operation
in Line~\ref{line:session2} takes a channel that is ready to receive a
value of type \lstinline{'a} and returns a pair of the value a the
channel at its residual type \lstinline{'S}. It does not matter
whether \lstinline{'a} is restricted to be linear, in fact
\lstinline{receive} is polymorphic in the kind of \lstinline{'a}.
This kind polymorphism is the
default if no constraints are specified.
%
The \lstinline{send} operation takes a linear channel and returns a
single-use function that takes a value of type \lstinline{'a} suitable
for sending and returns the channel with updated type.
%
The \lstinline{create} operation returns a pair of channel
endpoints. They follow dual communication protocols, where the
\lstinline{dual} operator swaps sending and receiving operations.
%
Finally, \lstinline{close} closes the channel.

@@

\cref{ex:array:region} uses braces as in \lstinline/{| ... |}/ to
explicitly state the regions of the example in \cref{ex:array}.  One
region consists of the header expression of the \lstinline/let/ in
\cref{ex:array:region:get}. It is contained in another region that
comprises the body of the \lstinline/for/ loop. \lang guarantees that borrows 
never escape the smallest enclosing region. It employs a system of
\emph{indexed kinds} like \lstinline{aff_r} and \lstinline{un_r} where
$r$ is a positive integer that corresponds to the lexical nesting
depth of regions. For instance, the type of \lstinline{&!a} in
\cref{ex:array:set} has kind \lstinline{aff_1} whereas the type of
\lstinline{&a} in \cref{ex:array:region:get} has kind
\lstinline{un_2} and the typing of its region is such that types with
kind indexes greater than or equal to $2$ cannot escape.

@@

\cref{sig:hybarray} contains an API \lstinline/HybArray/ 
that enables using mutable and immutable modifications to
the board through affine types and borrows.
The signature differs slightly from the \lstinline/Array/
signature. For simplicity, the array elements are unrestricted, but
the structure itself remains linear so as to be implemented in terms of \lstinline/Array/.
The in-place mutation function \lstinline/set_mut/ with type
\lstinline/&!('a t) * int * 'a -> unit/ works on a mutable borrow  whereas the persistent
\lstinline/set/ operation has type \lstinline/&('a t) * int * 'a -> 'a t/. It
takes an immutable borrow because it only reads from the
argument array, but returns a fresh,  modified structure.
The code in \cref{ex:cow} contains a very simple implementation of
\lstinline/HybArray/ that represents hybrid arrays
as regular arrays and uses copy-on-write for persistent
modifications. The function
\lstinline/mapi : (int -> &'a -> 'b) -> &('a t) -> 'b t /
is a simple variation on \lstinline/Array.map/ where the mapping function
takes the position of the element.
% A more elaborate implementation could use persistent arrays
% \cite{DBLP:conf/ml/ConchonF07}.

@@

  let set (a, i0, x0) =
    let f i x =
      if i = i0 then x0 else x
    in
    Array.mapi f a

@@

The board is represented as a 2D-matrix (\cref{line:boardtype}), where
the \lstinline/Matrix/ type uses the same API as \lstinline/HybArray/
but with two indices. 
Each cell contains an integer set that represent admissible solutions so far.
The Sudoku solver iterates over the cells and tries each possible solution (\cref{line:try_solution}).
When a value is picked for the current cell, we create a choice point,
change the cell with an immutable modification (\cref{line:immutset}), and propagate
the changes with the \lstinline/propagate/ function.
The \lstinline/propagate/ function uses direct mutation through a
mutable borrow of the matrix as it need not preserve the previous
version of the board.  
The implementation of \lstinline/propagate/ is split into three parts
for lines, columns, and square, which are all very similar to function
\lstinline/propagate_lines/ (\cref{line:propline}).

The typing ensures that the mutations do not compromise the state at
the choice point, because they operate on a new state \lstinline/new_g/ created for one
particular branch of the choice.
In particular, the \lstinline/try_solution/ function can be used in a loop
precisely because the \lstinline/set/ function only requires an
unrestricted immutable borrow of \lstinline/g/, which can be used as
many times as needed.
This would also easily allow parallel iteration on \cref{line:iter}.
As \lstinline/g/ is linear we must free it in \cref{line:free:g}.

@@

The implementation in \cref{impl:pool} represents a bag of resources
using a concurrent queue with atomic add and remove operations. 
The implementation of the \lstinline/Pool.create/ and \lstinline/Pool.consume/
functions is straightforward.
The function \lstinline/Pool.use/ first draws
an element from the pool (or creates a fresh element),
passes it down to the callback function \lstinline/f/, and returns
it to the pool afterwards.
For clarity,
we explicitly delimit the region in \cref{line:pool:region} to ensure that
the return value of \lstinline/f &!o/ does not capture \lstinline/&!o/.
In practice, such annotation is unnecessary, as it would be automatically
annotated.

\begin{figure}[tp]
  \centering
  \begin{subfigure}[t]{.58\linewidth}
\begin{lstlisting}
type ('a:'k) pool : 'k

@@

let uri = "postgresql://localhost:5432"
let main users =
  (* Create a database connection pool. *)
  let pool = Pool.create (fun _ -> Db.connect uri) in
  List.parallel_iter (create_user &pool) users;
  Pool.consume (fun c -> Db.close c)
\end{lstlisting}
    \caption{Example of use}
    \label{ex:pool}
  \end{subfigure}~
  \begin{subfigure}[t]{.43\linewidth}
\begin{lstlisting}
type ('a:'k) pool : 'k = {
  spawn : unit -> 'a ;
  queue : 'a CQueue.t;
}
